Binocular integration of chromatic and luminance signals

### 2024-04-18 manuscript.tex
Update to threshold criteria

\begin{equation}
Stage1_L = \frac{C_L^m}{S + C_L + \omega C_R},
\end{equation}

\begin{equation}
Stage1_R = \frac{C_R^m}{S + C_R + \omega C_L},
\end{equation}

\begin{equation}
binsum = Stage1_L + Stage1_R,
\end{equation}

\begin{equation}
Stage2 = \frac{binsum^p}{Z + binsum^q},
\end{equation}

@@

\begin{equation}
Stage2_{target+pedestal} - Stage2_{pedestal} = k,
\end{equation}

and d-prime for a single target level is defined as:

\begin{equation}
d' = \frac{Stage2_{target+pedestal} - Stage2_{pedestal}}{k}.
\end{equation}

@@

\begin{equation}
ACStage1_L = \frac{AC_L^m}{S + AC_L + \omega_A AC_R + \omega_R RG_R + \omega_B BY_R},
\end{equation}

@@

\centering
\begin{tabular}{lrrrrrrrr}
\toprule
\textbf{\em{Model fit}} & \textbf{\em{p}} & \textbf{\em{q}} & \textbf{\em{m}} & \textbf{\em{S}} & \textbf{\em{Z}} & \textbf{\em{$\omega$}} & \textbf{\em{k}} & \textbf{\em{RMSE}}\\
\midrule
Meese et al. (2006) & 7.99 & 6.59 & 1.28 & 0.99 & 0.08 & 1.00 & 0.19 & \\
\midrule
\textbf{Simplex fits} & \textbf{} & \textbf{} & \textbf{} & \textbf{} & \textbf{} & \textbf{} & \textbf{} & \textbf{}\\
Achromatic gratings & 5.96 & 4.69 & 1.34 & 0.87 & 0.13 & 1.00 & 0.20 & 1.59 dB\\
Red/green gratings & 7.77 & 5.19 & 1.17 & 0.55 & 0.07 & 1.00 & 0.17 & 1.39 dB\\
Blue/yellow gratings & 16.01 & 11.56 & 1.13 & 0.33 & 0.00 & 0.99 & 0.27 & 1.46 dB\\
Flickering discs & 5.62 & 4.51 & 1.26 & 0.95 & 0.25 & 0.89 & 0.13 & 1.05 dB\\
\midrule
\textbf{Bayesian model} & \textbf{} & \textbf{} & \textbf{} & \textbf{} & \textbf{} & \textbf{} & \textbf{} & \textbf{}\\
Achromatic gratings & 7.12 & 5.79 & 1.27 & 0.65 & 0.07 & 0.99 & 0.31 & \\
Red/green gratings & 5.25 & 3.60 & 1.23 & 1.27 & 0.03 & 1.05 & 0.20 & \\
Blue/yellow gratings & 6.78 & 5.17 & 1.24 & 0.58 & 0.01 & 0.96 & 0.36 & \\
Flickering discs & 7.21 & 5.75 & 1.25 & 0.69 & 0.13 & 0.88 & 0.26 & \\
\bottomrule
\end{tabular}
\end{table}

For consistency with previous work, we initially performed least-squares
fits of the two-stage model (Meese et al., 2006) for each of the
chromaticity experiments, and the flickering disc experiment (7 free
parameters per fit). The best model fits are shown by the curves in
Figure~\ref{fig-dipperfig} and Figure~\ref{fig-discdata}, and provide an
excellent description of the data, with RMS errors between 1.05 and
1.59dB. Best fitting parameters are shown in the `Simplex fits' section
of Table~\ref{tbl-parametertable}. We note that in previous work, the
weight of interocular suppression (\(\omega\) in the model) was
implicitly fixed at 1. Here we allowed it to vary, but it still received
a value very close to 1 in all of our grating conditions, and slightly
below 1 (0.89) for our flickering discs. The exponent values (\emph{p}
and \emph{q}) for the blue/yellow gratings are rather different from
those of the other conditions. In previous work (Meese et al., 2006) the
second stage gain control nonlinearity (Equation 4) does typically have
quite substantial exponent values, which balance the relatively mild
nonlinearity at the first stage, and produce a compressive transducer
that results in contrast masking (the handle of the dipper). The high
value of \emph{p} = 16.01 is therefore likely to be compensating for the
low value of \emph{m} = 1.13 at stage one, but may well represent the
combination of several successive stages of nonlinearity, and perhaps
also other phenomena such as uncertainty (Pelli, 1985).

@@

Estimates of the binocular summation ratio at threshold fell within the
range \(\sqrt{2}\) to 2 for all four stimuli tested here. Consistent
with previous reports (Simmons, 2005), summation was slightly higher for
chromatic versus achromatic stimuli, but fell short of perfect linear
summation (a ratio of 2). In our model, summation at threshold is
determined by the exponent at the first gain control stage (the \emph{m}
parameter in Equations 1 \& 2). The summation ratio can be approximated
by \(2^{1/m}\) (see Figure 9 of Baker et al., 2012), such that an
exponent of \(m=1\) produces a ratio of 2, and an exponent of \(m = 2\)
produces a ratio of \(\sqrt{2}\). Both the simplex and Bayesian model
fits (see Table~\ref{tbl-parametertable}) generated parameter estimates
in the range \(1 < m < 1.4\), consistent with the high levels of
summation observed empirically.

@@

The weight of interocular suppression is represented by the model
parameter, \(\omega\) (Equations 1 \& 2). Posterior distributions of
this parameter overlapped 1 for all of our dipper function data sets
(see top right panel in Figure~\ref{fig-bayesianmodel}), indicating
strong interocular suppression regardless of post-retinal pathway. The
flickering disc stimuli produced a slightly lower suppression estimate,
with \(\omega \sim 0.9\). This may indicate slightly weaker suppression
between the eyes for temporal modulations, but it is much less extreme
than our recent estimates using EEG and matching paradigms (Segala et
al., 2023). One difference between these studies is the luminance of the
background, which was set to black for the experiments of Segala et al.
(2023), but here was set at the mean luminance. Other studies using
static luminance increments have also reported differences in the
character of binocular combination that are attributable to background
luminance (Baker et al., 2012), so this may explain the differences
between studies. We also note that the weaker interocular suppression
for flickering disc stimuli appears to be largely due to one participant
(see Figure~\ref{fig-individualdiscs}c), so individual differences might
also play a role. Future work should manipulate the background luminance
systematically to better understand how this modulates interocular
suppression.

@@

Recent evidence indicates that the physiological substrate of
interocular suppression may be neurons in layer 4 of primary visual
cortex (Dougherty et al., 2019). Most cells in this layer are
monocularly excitable, in that their responses increase only by
stimulation of their preferred eye. However, simultaneous stimulation of
the non-preferred eye can modulate the response, usually in an
inhibitory fashion, exactly as proposed at stage 1 of the two-stage
model (Equations 1 \& 2). In terms of perception, one consequence of
this early suppression is to achieve `ocularity invariance', whereby the
perceived contrast of a stimulus viewed by one eye is equivalent to that
of the same stimulus viewed by both eyes (Baker et al., 2007). Similar
processes of response invariance have also been reported using fMRI
(Moradi \& Heeger, 2009) and steady-state EEG (Baker \& Wade, 2017).

>>>>

\begin{equation}\protect\hypertarget{eq-s1l}{}{\mathrm{Stage1}_L = \frac{C_L^m}{S + C_L + \omega C_R},}\label{eq-s1l}\end{equation}

\begin{equation}\protect\hypertarget{eq-s1r}{}{\mathrm{Stage1}_R = \frac{C_R^m}{S + C_R + \omega C_L},}\label{eq-s1r}\end{equation}

\begin{equation}\protect\hypertarget{eq-binsum}{}{\mathrm{binsum} = \mathrm{Stage1}_L + \mathrm{Stage1}_R,}\label{eq-binsum}\end{equation}

\begin{equation}\protect\hypertarget{eq-s2}{}{\mathrm{Stage2} = \frac{\mathrm{binsum}^p}{Z + \mathrm{binsum}^q},}\label{eq-s2}\end{equation}

@@

\begin{equation}\protect\hypertarget{eq-thd}{}{\mathrm{Stage2}_{\mathrm{target+pedestal}} - \mathrm{Stage2}_{\mathrm{pedestal}} = k,}\label{eq-thd}\end{equation}

and \(d'\) (d-prime) for a single target level is defined as:

\begin{equation}\protect\hypertarget{eq-dprime}{}{d' = \frac{\mathrm{Stage2}_{\mathrm{target+pedestal}} - \mathrm{Stage2}_{\mathrm{pedestal}}}{k/0.954},}\label{eq-dprime}\end{equation}

where the constant 0.954 reflects the value of \(d'\) at the 75\%
correct point on the psychometric function used to estimate threshold.

@@

\begin{equation}\protect\hypertarget{eq-s1multi}{}{^{AC}\mathrm{Stage1}_L = \frac{AC_L^m}{S + AC_L + \omega_A AC_R + \omega_R RG_R + \omega_B BY_R},}\label{eq-s1multi}\end{equation}

@@

\centering
\begin{tabular}{lrrrrrrrr}
\toprule
\textbf{\em{Model fit}} & \textbf{\em{p}} & \textbf{\em{q}} & \textbf{\em{m}} & \textbf{\em{S}} & \textbf{\em{Z}} & \textbf{\em{$\omega$}} & \textbf{\em{k}} & \textbf{\em{RMSE}}\\
\midrule
Meese et al. (2006) & 7.99 & 6.59 & 1.28 & 0.99 & 0.08 & 1.00 & 0.19 & \\
\midrule
\textbf{Simplex fits} & \textbf{} & \textbf{} & \textbf{} & \textbf{} & \textbf{} & \textbf{} & \textbf{} & \textbf{}\\
Achromatic gratings & 5.96 & 4.69 & 1.34 & 0.87 & 0.13 & 1.00 & 0.20 & 1.59 dB\\
Red/green gratings & 7.77 & 5.19 & 1.17 & 0.55 & 0.07 & 1.00 & 0.17 & 1.39 dB\\
Blue/yellow gratings & 16.01 & 11.56 & 1.13 & 0.33 & 0.00 & 0.99 & 0.27 & 1.46 dB\\
Flickering discs & 5.62 & 4.51 & 1.26 & 0.95 & 0.25 & 0.89 & 0.13 & 1.05 dB\\
\midrule
\textbf{Bayesian model} & \textbf{} & \textbf{} & \textbf{} & \textbf{} & \textbf{} & \textbf{} & \textbf{} & \textbf{}\\
Achromatic gratings & 7.06 & 5.63 & 1.27 & 0.63 & 0.08 & 0.98 & 0.30 & \\
Red/green gratings & 5.11 & 3.63 & 1.24 & 1.22 & 0.03 & 1.05 & 0.19 & \\
Blue/yellow gratings & 6.78 & 5.34 & 1.24 & 0.61 & 0.01 & 0.95 & 0.34 & \\
Flickering discs & 7.16 & 5.76 & 1.25 & 0.65 & 0.13 & 0.87 & 0.24 & \\
\bottomrule
\end{tabular}
\end{table}

For consistency with previous work, we initially performed least-squares
fits of the two-stage model (Meese et al., 2006) for each of the
chromaticity experiments, and the flickering disc experiment (7 free
parameters per fit). The best model fits are shown by the curves in
Figure~\ref{fig-dipperfig} and Figure~\ref{fig-discdata}, and provide an
excellent description of the data, with RMS errors between 1.05 and
1.59dB. Best fitting parameters are shown in the `Simplex fits' section
of Table~\ref{tbl-parametertable}. We note that in previous work, the
weight of interocular suppression (\(\omega\) in the model) was
implicitly fixed at 1. Here we allowed it to vary, but it still received
a value very close to 1 in all of our grating conditions, and slightly
below 1 (0.89) for our flickering discs. The exponent values (\emph{p}
and \emph{q}) for the blue/yellow gratings are rather different from
those of the other conditions. In previous work (Meese et al., 2006) the
second stage gain control nonlinearity (Equation~\ref{eq-s2}) does
typically have quite substantial exponent values, which balance the
relatively mild nonlinearity at the first stage, and produce a
compressive transducer that results in contrast masking (the handle of
the dipper). The high value of \emph{p} = 16.01 is therefore likely to
be compensating for the low value of \emph{m} = 1.13 at stage one, but
may well represent the combination of several successive stages of
nonlinearity, and perhaps also other phenomena such as uncertainty
(Pelli, 1985).

@@

Estimates of the binocular summation ratio at threshold fell within the
range \(\sqrt{2}\) to 2 for all four stimuli tested here. Consistent
with previous reports (Simmons, 2005), summation was slightly higher for
chromatic versus achromatic stimuli, but fell short of perfect linear
summation (a ratio of 2). In our model, summation at threshold is
determined by the exponent at the first gain control stage (the \emph{m}
parameter in Equation~\ref{eq-s1l} and Equation~\ref{eq-s1r}). The
summation ratio can be approximated by \(2^{1/m}\) (see Figure 9 of
Baker et al., 2012), such that an exponent of \(m=1\) produces a ratio
of 2, and an exponent of \(m = 2\) produces a ratio of \(\sqrt{2}\).
Both the simplex and Bayesian model fits (see
Table~\ref{tbl-parametertable}) generated parameter estimates in the
range \(1 < m < 1.4\), consistent with the high levels of summation
observed empirically.

@@

The weight of interocular suppression is represented by the model
parameter, \(\omega\) (Equation~\ref{eq-s1l} and Equation~\ref{eq-s1r}).
Posterior distributions of this parameter overlapped 1 for all of our
dipper function data sets (see top right panel in
Figure~\ref{fig-bayesianmodel}), indicating strong interocular
suppression regardless of post-retinal pathway. The flickering disc
stimuli produced a slightly lower suppression estimate, with
\(\omega \sim 0.9\). This may indicate slightly weaker suppression
between the eyes for temporal modulations, but it is much less extreme
than our recent estimates using EEG and matching paradigms (Segala et
al., 2023). One difference between these studies is the luminance of the
background, which was set to black for the experiments of Segala et al.
(2023), but here was set at the mean luminance. Other studies using
static luminance increments have also reported differences in the
character of binocular combination that are attributable to background
luminance (Baker et al., 2012), so this may explain the differences
between studies. We also note that the weaker interocular suppression
for flickering disc stimuli appears to be largely due to one participant
(see Figure~\ref{fig-individualdiscs}c), so individual differences might
also play a role. Future work should manipulate the background luminance
systematically to better understand how this modulates interocular
suppression.

@@

Recent evidence indicates that the physiological substrate of
interocular suppression may be neurons in layer 4 of primary visual
cortex (Dougherty et al., 2019). Most cells in this layer are
monocularly excitable, in that their responses increase only by
stimulation of their preferred eye. However, simultaneous stimulation of
the non-preferred eye can modulate the response, usually in an
inhibitory fashion, exactly as proposed at stage 1 of the two-stage
model (Equation~\ref{eq-s1l} and Equation~\ref{eq-s1r}). In terms of
perception, one consequence of this early suppression is to achieve
`ocularity invariance', whereby the perceived contrast of a stimulus
viewed by one eye is equivalent to that of the same stimulus viewed by
both eyes (Baker et al., 2007). Similar processes of response invariance
have also been reported using fMRI (Moradi \& Heeger, 2009) and
steady-state EEG (Baker \& Wade, 2017).
